---
title: "The Leslie Matrix - revisited"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are told that annually:

-   Each breeding adult will produce, on average, 600 eggs
-   5% of eggs will survive to become fry
-   20% of fry will survive to become juveniles
-   16% of juveniles will survive to become adults
-   Adult survival rate year to year is 90%

Recall the Leslie matrix from EDS 212. Matrix multiplication (using the dot product `%*%` of the matrix with the populations at each life stage) yields the population structure at the following year.

### Create the Leslie matrix:

```{r}
# Make the Leslie Matrix:
fish_leslie <- matrix(c(0, 0, 0, 600, 0.05, 0, 0, 0, 0, 0.20, 0, 0, 0, 0, 0.16, 0.9), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out: 
fish_leslie
```

The initial population structure is given by:

```{r}
# Initial population structure (Roe, Fry, Juvenile, Adult): 
fish_year0 <- c(5000, 8000, 600, 400)
```

-   take the dot product of the Leslie matrix and the initial population structure to get the following year's population structure

-   dot product of matrix and population structure at i-th time cycle

Add line-by-line comments wheres prompted in the code below, explaining what each is doing. Make sure that you are talking with your neighbors about what is really being created at each step.

Notes from class:

-   propagation ...?

-   initial pop structure & conditions state that those proportions are the only way that eggs, fry, etc. exist -- also based on the previous year's population

```{r}
# Model projections

# ------ WHAT HAPPENS HERE? ------ #
# ------ comment line-by-line ---- #

# The line below creates a vector of time in years over the span of 8 years, including our initial state (0 years)
time_yr <- seq(from = 0, to = 8, by = 1)
# time_yr returns a sequence from 0 to 8 in increments of 1


# The line below establishes an empty / storage matrix with the length of the vector of initial population conditions (roe, fry, juvenille, adult) as the number of columns, with the length of the time sequence of years as the number of rows. 
# we can always check individual lengths of elements with length(time_yr) etc. 
proj_year <- matrix(ncol = length(fish_year0), nrow = length(time_yr))


# The line below is filling the empty matrix with the first row as the initial population conditions (because that's year 0!)
proj_year[1, ] <- fish_year0


# The for loop below starts after year 0 (the 1st position/element in the time_yr vector) because we already have our initial population structure. This for loop then iterates through the years, populating the proj_year matrix by row, given the starting conditions, then the population structure for the previous year (i - 1) 
# proj_year[i-1,] returns the vector containing the population structure data of the previous (time) cycle (previous row to the one you're filling)
# each iteration populates a new row

for (i in 2:length(time_yr)) {
  proj_year[i,] <- fish_leslie %*% proj_year[i-1,]
}


# The rest is wrangling & visualization (run to check it out):
# reassigns the column names of the matrix to be more meaningful, also useful in plotting 
colnames(proj_year) <- c("roe", "fry", "juvenile", "adult")

# binding together the values in our matrix with our time_yr vector into a data frame (now we have column names, and years as column 1 values)
proj_df <- data.frame(time_yr, proj_year) %>% 
  # excludes time_yr column -- we could also have written cols = c("eggs", "fry", "juvenille", "adult")
  # values given the name "stage population"
  pivot_longer(cols = -time_yr, names_to = "lifestage", values_to = "stage_population")

ggplot(data = proj_df, aes(x = time_yr, y = stage_population)) +
  geom_line(aes(color = lifestage)) +
  # adds log scaling because the scales are very different
  scale_y_log10()

```

Test different combinations of survival / reproduction and starting population structures to see what happens!

```{r}
demo <- data.frame(year = c(1995, 2000, 2005), 
                  cats = c(2, 4, 6),
                  dogs = c(10, 14, 12))

demo
```

```{r}
ggplot(data = demo) +
  geom_line(aes(x = year, y = cats)) +
  geom_line(aes(x = year, y = dogs))

# this is more tedious than asking it to recognize multiple groups as single variables

# we want a single column for each variable
```

```{r}
demo_longer <- demo |> 
  # take column names, put them into one value
  # name the column that they get melted into 
  # name what the values column is
  pivot_longer(cols = c(cats, dogs), names_to = "species", values_to = "count")
  
demo_longer
```

## Testing out sourcing scripts

```{r}
source(here::here("src", "name_cart.R"))


name_cart(food = "sushi", animal = "tuna")
```
